### 2022-08-23 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Second round of internal revision.

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations whose
second derivative functions often have large local $L^2$ norm
(approximations with large {\it wiggle}). In an attempt to capture the
local shape of the data while minimizing wiggle, this package uses a
facet model from image processing [Haralick and Watson 1981] to
estimate first and second derivatives at breakpoints. Rather than
picking a local linear or quadratic fit with minimal residual, this
work uses a quadratic facet model that selects the local quadratic
interpolant with minimum magnitude second derivative.

@@

\nobreak
where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
\le n$, and $n \ge 3$. Returns the first and second derivative at 
$X_i$ of the local quadratic interpolant with minimum magnitude
second derivative. Approximate equality is denoted with $\approx$ and
states that two numbers are considered equal once they are within the
machine precision $\epsilon$ of each other.

@@

\nobreak
where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$. Approximate equality is denoted with
$\approx$ and states that two numbers are considered equal once they
are within the machine precision $\epsilon$ of each other.

@@

Given the minimum curvature (minimum magnitude of the second derivative)
nature of the initial derivative estimates, it is desirable to make
the smallest necessary changes to the initial interpolating spline $Q$
while enforcing monotonicity. In practice a binary search for the
boundary of monotonicity is used in place of solely shrinking $DQ$ and
$D^2Q$ at breakpoints adjoining {\it active\/} intervals: intervals
over which $Q$ is nonmonotone at least once during the search. The
binary search considers a Boolean function $B_i(s)$, for $0 \le s \le
1$, that is true if the order six polynomial piece of $Q(x)$ on
$[X_i, X_{i+1}]$ matching derivatives
$Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false
otherwise.  The binary search is only applied at those breakpoints
adjoining intervals $[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and
hence $B_i(1)$ is false.  It is further assumed that there exists
$0 \le s^* \le 1$ such that $B_i(s)$ is true for $0 \le s \le s^*$ and
false for some $1 > s > s^*$. Since the derivative conditions at
interior breakpoints are shared by intervals left and right of the
breakpoint, the binary search is performed at all breakpoints
simultaneously.  Specifically, the monotonicity of $Q$ is checked on
all active intervals in each step of the binary search to determine
the next derivative modification at each breakpoint. The goal of this
search is to converge on the boundary of the monotone region in the
$(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
Watson [1994]) for all intervals. This multiple-interval binary search
allows the value zero to be obtained for all (first and second)
derivative values in a fixed maximum number of computations, hence has
no effect on computational complexity order. This binary search
algorithm is outlined below.

>>>>

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations whose
second derivative functions often have large local $L^2$ norm
(approximations with large {\it wiggle}). In an attempt to capture the
local shape of the data while minimizing wiggle, this package uses a
facet model from image processing [Haralick and Watson 1981] to
estimate first and second derivatives at breakpoints. Rather than
picking a local linear or quadratic fit with minimal residual, this
work \red{instead} % uses a quadratic facet model that
selects the local quadratic
interpolant with minimum magnitude second derivative \red{(hereafter
referred to as a {\it quadratic facet model}). The term ``facet'' 
refers to one piece of an image surface modelled by a two-dimensional
piecewise polynomial.}.

@@

\nobreak
where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
\le n$, and $n \ge 3$. Returns the first and second derivative at 
$X_i$ of the local quadratic interpolant with minimum magnitude
second derivative. Approximate equality is denoted with
$\approx$ and considers two numbers to be equal once they are within
the machine precision $\epsilon$ of each other.

@@

\nobreak

where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$. Approximate equality is denoted with
$\approx$ and considers two numbers to be equal once they are within
the machine precision $\epsilon$ of each other.

@@

Given \red{that} the \red{initial derivative estimates locally minimize 
the magnitude of the second derivative}, it is desirable to make
the smallest necessary changes to the initial interpolating spline $Q$
while enforcing monotonicity. In practice a binary search for the
boundary of monotonicity is used in place of solely shrinking $DQ$ and
$D^2Q$ at breakpoints adjoining {\it active\/} intervals: intervals
over which $Q$ is nonmonotone at least once during the search. The
binary search considers a Boolean function $B_i(s)$, for $0 \le s \le
1$, that is true if the order six polynomial piece of $Q(x)$ on
$[X_i, X_{i+1}]$ matching derivatives
$Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false
otherwise.  The binary search is only applied at those breakpoints
adjoining intervals $[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and
hence $B_i(1)$ is false.  It is further assumed that there exists
$0 \le s^* \le 1$ such that $B_i(s)$ is true for $0 \le s \le s^*$ and
false for some $1 > s > s^*$. Since the derivative conditions at
interior breakpoints are shared by intervals left and right of the
breakpoint, the binary search is performed at all breakpoints
simultaneously.  Specifically, the monotonicity of $Q$ is checked on
all active intervals in each step of the binary search to determine
the next derivative modification at each breakpoint. The goal of this
search is to converge on the boundary of the monotone region in the
$(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
Watson [1994]) for all intervals. This multiple-interval binary search
allows the value zero to be obtained for all (first and second)
derivative values in a fixed maximum number of computations, hence has
no effect on computational complexity order. This binary search
algorithm is outlined below.

### 2022-08-23 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Second round of internal revision.

% Abstract and Introduction
{\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
monotone quintic spline interpolants to \red{univariate} monotone data. Using sharp
theoretical monotonicity constraints, first and second derivative estimates
at data provided by a quadratic facet model are refined to produce a
\red{univariate} $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
details, complexity and sensitivity analyses, usage information, a
brief performance study, \red{and comparisons with other spline approaches} are included.

@@

%% No extra knots, rational functions, sufficiency and necessity
%% proven, works for unequally spaced data, no code associated with
%% the paper, no algorithm provided for the initialization for free
%% shape parameters.
\refj{\beginred Abbas M., Majid, A. A. and Ali,
  J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for
  monotone data}{Applied Mathematics and
  Computation}{219(6):2885--2895}
\endred

@@

%% Proves existence of splines that maintain monotonicity and
%% convexity while having fixed continuity and order.
\refj{\beginred Costantini, P.}{1986}{On monotone and convex spline
  interpolation}{Mathematics of Computation}{46(173):203--214}
\endred

%% Source paper accompanying the BVSPIS publication.
\refj{\beginred Costantini, P.}{1997a}{Boundary-valued shape-preserving
  interpolating splines}{ACM Transactions on Mathematical Software
  (TOMS)}{23(2):229--251}
\endred

%% BVSPIS Algorithm.
\refj{\beginred Costantini, P.}{1997b}{Algorithm 770: BVSPIS -- A package for
  computing boundary-valued shape-preserving interpolating
  splines}{ACM Transactions on Mathematical Software
  (TOMS)}{23(2):252-–254}
\endred

%% Constructs C3 interpolating splines by progressively increasing
%% tension through Bezier control polygons until shape parameters are
%% satisfied, no code is provided.
\refj{\beginred Cravero, I. and C. Manni}{2003}{Shape-preserving interpolants with
  high smoothness}{Journal of computational and applied
  mathematics}{157(2):383--405}
\endred

%% Iterates towards a solution of nonlinear equations that ensure C2
%% continuity of quadratic rational splines. The solution they iterate
%% towards exists and is unique. They provide proofs and methods for
%% setting derivatives at the ends, no formal algorithms or code.
\refj{\beginred Delbourgo, R. and Gregory, J. A.}{1983}{$C^2$ rational quadratic
  spline interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{3(2):141--152}
\endred

%% Continuous previous work by incorporating a cubic numerator, proves
%% shape preservation when the correct value for a tension parameter
%% is set but no method for setting that parameter. Accuracy depends
%% on accurate derivative evaluations. No code nor formal algorithm is
%% provided.
\refj{\beginred Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
  tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}
\endred

%% Proves approximation error bounds for monotone spline interpolation
%% of data from monotone functions.
\refj{\beginred DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
  Journal on Mathematical Analysis}{8(5):891--905}
\endred

%% Proposes an algorithm for constructing C2 Hermite monotone spline
%% interpolants using proven sufficient, but not necessary
%% conditions. As these bounds are not sharp, the space of allowable
%% derivative and second derivative values at knots is unnecessarily
%% shrunk. An algorithm is provided, but no code. Recommends against
%% the direct optimization for smoothness parameters like L2, instead
%% promoting problem specific definitions of 'geometric niceness'.
\refj{\beginred Dougherty, R. L., Edelman, A. and Hyman,
  J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving
  cubic and quintic Hermite interpolation}{Mathematics of
  Computation}{52(186):471--494}
\endred

%% Proves a simple method for determining the existence of a monotone
%% C2 cubic interpolating spline, but functions of this kind do not
%% always exist for arbitrary convex monotone data sets.
\refj{\beginred Fiorot, J. C. and Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
  polynomial interpolating splines}{Mathematics of
  Computation}{57(195):291--298} \endred \goodbreak

@@

%% Proves necessary and sufficient conditions for a closed form
%% solution to monotone C1 quadratic rational spline
%% interpolation. Does not provide formal algorithm, nor code.
\refj{\beginred Gregory, J. A. and Delbourgo, R.}{1982}{Piecewise rational
  quadratic interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{2(2):123--130}
\endred

@@

% The paper where a linear facet model is proposed.
\refj{\beginred Haralick, R. M. and Watson, L. T.}{1981}{A facet model for image
  data}{Computer Graphics Image Processing}{15:113--129}
\endred

%% Proposes a method for making arbitrarily smooth shape preserving
%% spline interpolants, but uses a centered quadratic to estimate
%% derivatives and a sufficient set of bounding conditions to ensure
%% shape preservation in practice. No necessary conditions are
%% applied, no algorithm is provided, neither is any code provided.
\refj{\beginred Han, X.}{2018}{Shape-preserving piecewise rational interpolation
  with higher order continuity}{Applied Mathematics and
  Computation}{337:1--13}
\endred

@@

%% Presents several nonlinear boundary conditions based on the median
%% function to establish a fourth order monotonicity preserving
%% constraints relying on divided difference initial derivative
%% estimates. Overall generates a monotone C1 cubic interpolant to
%% data.
\refj{\beginred Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
  Journal of Numerical Analysis}{30(1):57--100}
\endred

@@

% Monotone quintic algorithm.
\refj{\beginred Lux, T. C. H., Watson, L. T., Chang, T. H., Xu, L.,
  Wang, Y. and Hong, Y.}{2020}{An algorithm for constructing monotone
  quintic interpolating splines}{SCS {\sl Spring Simulation
    Multiconference, High Performance Computing Symposium}}{Article
  33, 1--12}
\endred

%% Constructs a C2 cubic spline with two additional knots per
%% interval. No code (paywall).
\refj{\beginred Manni, C. and Sablonni\`ere, P.}{1997}{Monotone interpolation of
  order 3 by $C^2$ cubic splines}{IMA Journal of Numerical
  Analysis}{17(2):305--320}
\endred

%% Uses sufficient conditions to generate C2 cubic spline. No code
%% (paywall).
\refj{\beginred Manni, C.}{2001}{On shape preserving $C^2$ Hermite
  interpolation}{BIT Numerical Mathematics}{41(1):127--148}
\endred

@@

% Proves that C2 convexity preserving splines do not always exist.
\reft{\beginred Mulansky, B. and Neamtu, M.}{1991}{On the existence of shape
  preserving interpolation operators}{Dissertation, University of
  Twente, Vancouver}
\endred

@@

%% Proposes a cubic C2 shape preserving spline method with free
%% parameters (mentions quintic splines are necessary for globally C2
%% monotone splines). Sufficient (but not necessary) conditions are
%% provided for monotonicity.
\refj{\beginred Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
  interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}
\endred

@@

% ulrich1990positivity tech report
\reft{\beginred Ulrich, G. and Watson, L. T.}{1990}{Positivity conditions for
  quartic polynomials}{Technical Report 90-57, Department of Computer
  Science, Virginia Polytechnic Institute and State University,
  Blacksburg, Virginia}
\endred

@@

%% Uses sufficient conditions over three free parameters to construct
%% C2 rational quartic shape preserving splines.
\refj{\beginred Zhu, Y. and Han, X.}{2015a}{$C^2$ rational quartic interpolation
  spline with local shape preserving property}{Applied Mathematics
  Letters}{46:57--63}
\endred

%% Uses sufficient conditions over two free parameters to construct C2
%% rational quartic shape preserving splines.
\refj{\beginred Zhu, Y. and Han, X.}{2015b}{Shape preserving $C^2$ rational quartic
  interpolation spline with two parameters}{International Journal of
  Computer Mathematics}{92(10):2160--2177}
\endred
}
%% -------------------------------------------------------------------

>>>>

% Abstract and Introduction
{\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
monotone quintic spline interpolants to univariate monotone data. Using sharp
theoretical monotonicity constraints, first and second derivative estimates
at data provided by a quadratic facet model are refined to produce a
univariate $\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
details, complexity and sensitivity analyses, usage information, a
brief performance study, and comparisons with other spline approaches are included.

@@

%% No extra knots, rational functions, sufficiency and necessity
%% proven, works for unequally spaced data, no code associated with
%% the paper, no algorithm provided for the initialization for free
%% shape parameters.
\refj{Abbas M., Majid, A. A. and Ali,
  J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for
  monotone data}{Applied Mathematics and
  Computation}{219(6):2885--2895}

@@

%% Proves existence of splines that maintain monotonicity and
%% convexity while having fixed continuity and order.
\refj{Costantini, P.}{1986}{On monotone and convex spline
  interpolation}{Mathematics of Computation}{46(173):203--214}

%% Source paper accompanying the BVSPIS publication.
\refj{Costantini, P.}{1997a}{Boundary-valued shape-preserving
  interpolating splines}{ACM Transactions on Mathematical Software
  (TOMS)}{23(2):229--251}

%% BVSPIS Algorithm.
\refj{Costantini, P.}{1997b}{Algorithm 770: BVSPIS -- A package for
  computing boundary-valued shape-preserving interpolating
  splines}{ACM Transactions on Mathematical Software
  (TOMS)}{23(2):252-–254}

%% Constructs C3 interpolating splines by progressively increasing
%% tension through Bezier control polygons until shape parameters are
%% satisfied, no code is provided.
\refj{Cravero, I. and C. Manni}{2003}{Shape-preserving interpolants with
  high smoothness}{Journal of computational and applied
  mathematics}{157(2):383--405}

%% Iterates towards a solution of nonlinear equations that ensure C2
%% continuity of quadratic rational splines. The solution they iterate
%% towards exists and is unique. They provide proofs and methods for
%% setting derivatives at the ends, no formal algorithms or code.
\refj{Delbourgo, R. and Gregory, J. A.}{1983}{$C^2$ rational quadratic
  spline interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{3(2):141--152}

%% Continuous previous work by incorporating a cubic numerator, proves
%% shape preservation when the correct value for a tension parameter
%% is set but no method for setting that parameter. Accuracy depends
%% on accurate derivative evaluations. No code nor formal algorithm is
%% provided.
\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
  tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}

%% Proves approximation error bounds for monotone spline interpolation
%% of data from monotone functions.
\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
  Journal on Mathematical Analysis}{8(5):891--905}

%% Proposes an algorithm for constructing C2 Hermite monotone spline
%% interpolants using proven sufficient, but not necessary
%% conditions. As these bounds are not sharp, the space of allowable
%% derivative and second derivative values at knots is unnecessarily
%% shrunk. An algorithm is provided, but no code. Recommends against
%% the direct optimization for smoothness parameters like L2, instead
%% promoting problem specific definitions of 'geometric niceness'.
\refj{Dougherty, R. L., Edelman, A. and Hyman,
  J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving
  cubic and quintic Hermite interpolation}{Mathematics of
  Computation}{52(186):471--494}

%% Proves a simple method for determining the existence of a monotone
%% C2 cubic interpolating spline, but functions of this kind do not
%% always exist for arbitrary convex monotone data sets.
\refj{Fiorot, J. C. and Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
  polynomial interpolating splines}{Mathematics of
  Computation}{57(195):291--298} \goodbreak

@@

%% Proves necessary and sufficient conditions for a closed form
%% solution to monotone C1 quadratic rational spline
%% interpolation. Does not provide formal algorithm, nor code.
\refj{Gregory, J. A. and Delbourgo, R.}{1982}{Piecewise rational
  quadratic interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{2(2):123--130}

@@

% The paper where a linear facet model is proposed.
\refj{Haralick, R. M. and Watson, L. T.}{1981}{A facet model for image
  data}{Computer Graphics Image Processing}{15:113--129}

%% Proposes a method for making arbitrarily smooth shape preserving
%% spline interpolants, but uses a centered quadratic to estimate
%% derivatives and a sufficient set of bounding conditions to ensure
%% shape preservation in practice. No necessary conditions are
%% applied, no algorithm is provided, neither is any code provided.
\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation
  with higher order continuity}{Applied Mathematics and
  Computation}{337:1--13}

@@

%% Presents several nonlinear boundary conditions based on the median
%% function to establish a fourth order monotonicity preserving
%% constraints relying on divided difference initial derivative
%% estimates. Overall generates a monotone C1 cubic interpolant to
%% data.
\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
  Journal of Numerical Analysis}{30(1):57--100}

@@

% Monotone quintic algorithm.
\refj{Lux, T. C. H., Watson, L. T., Chang, T. H., Xu, L.,
  Wang, Y. and Hong, Y.}{2020}{An algorithm for constructing monotone
  quintic interpolating splines}{SCS {\sl Spring Simulation
    Multiconference, High Performance Computing Symposium}}{Article
  33, 1--12}

%% Constructs a C2 cubic spline with two additional knots per
%% interval. No code (paywall).
\refj{Manni, C. and Sablonni\`ere, P.}{1997}{Monotone interpolation of
  order 3 by $C^2$ cubic splines}{IMA Journal of Numerical
  Analysis}{17(2):305--320}

%% Uses sufficient conditions to generate C2 cubic spline. No code
%% (paywall).
\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite
  interpolation}{BIT Numerical Mathematics}{41(1):127--148}

@@

% Proves that C2 convexity preserving splines do not always exist.
\reft{Mulansky, B. and Neamtu, M.}{1991}{On the existence of shape
  preserving interpolation operators}{Dissertation, University of
  Twente, Vancouver}

@@

%% Proposes a cubic C2 shape preserving spline method with free
%% parameters (mentions quintic splines are necessary for globally C2
%% monotone splines). Sufficient (but not necessary) conditions are
%% provided for monotonicity.
\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
  interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}

@@

% ulrich1990positivity tech report
\reft{Ulrich, G. and Watson, L. T.}{1990}{Positivity conditions for
  quartic polynomials}{Technical Report 90-57, Department of Computer
  Science, Virginia Polytechnic Institute and State University,
  Blacksburg, Virginia}

@@

%% Uses sufficient conditions over three free parameters to construct
%% C2 rational quartic shape preserving splines.
\refj{Zhu, Y. and Han, X.}{2015a}{$C^2$ rational quartic interpolation
  spline with local shape preserving property}{Applied Mathematics
  Letters}{46:57--63}

%% Uses sufficient conditions over two free parameters to construct C2
%% rational quartic shape preserving splines.
\refj{Zhu, Y. and Han, X.}{2015b}{Shape preserving $C^2$ rational quartic
  interpolation spline with two parameters}{International Journal of
  Computer Mathematics}{92(10):2160--2177}
}
%% -------------------------------------------------------------------

\input appendix
